Composable Concurrency Models

### 2016-09-29 paper/paper.tex
Panic-wrote section draft

\documentclass{sig-alternate}
\usepackage{biblatex}

\author{
\alignauthor{}
Dan Stelljes\\
  \affaddr{Division of Science and Mathematics}\\
  \affaddr{University of Minnesota, Morris}\\
  \affaddr{Morris, Minnesota, USA 56267}\\
  \email{stell124@morris.umn.edu}
}
\conferenceinfo{UMM CSci Senior Seminar Conference, December 2016}{Morris, MN}
\title{Composable Concurrency Abstractions}

@@

\begin{abstract}
Varied concurrency models and implementations present challenges in both performance and composability. By identifying common concepts and techniques, a high-level language virtual machine can support a variety of abstractions at a low level, and different models can be used in tandem.
\end{abstract}

\keywords{concurrency abstraction, concurrency, parallelism, virtual machines}

@@

Concurrency means running multiple tasks in overlapping time periods. It's desirable for a couple of important reasons: complex tasks can be completed faster by splitting them into simple parts that can run concurrently, and high-latency tasks can execute without blocking other tasks.

Concurrent tasks can be controlled in different ways (event loops, software transactional memory, etc.). Someone writing a program might like to mix different control mechanisms to take advantage of particular performance or semantic traits.

\section{Abstraction requirements}

Ziv et al.~\cite{Ziv2015} formalized desired requirements for concurrency abstractions to define a protocol for composition.

@@

A history of concurrent operations is serializable if its outcome is the same as the outcome of the operations executed consecutively.

\subsection{Deadlock avoidance}

Progress guarantees avoid deadlock, the state in which two concurrent operations are stuck waiting for the other to finish.

\subsection{Opacity}

A concurrency protocol is opaque if it ensures that each concurrent operation sees a consistent state.

\section{Runtime requirements}

Marr and D'Hondt~\cite{Marr2012} surveyed concurrent programming concepts with the goal of improving support at the virtual machine level. Such support could improve performance and enforce semantic guarantees between languages with a common virtual machine.

\subsection{Managed mutation and execution}

To ensure serializability, any modification of an object's state must be regulated.

\subsection{Ownership}

Mutation and execution must be regulated by some owning entity.

\subsection{Leveled reflection}

Safe language-level reflection and unsafe ``meta-level'' reflection need to be distinguished.

\subsection{Enforcability}

All rules need to be enforceable across all concurrency models.

>>>>

\documentclass{sig-alternate}
\usepackage{biblatex}
\usepackage{listings}

\newcommand{\code}[1]{\texttt{#1}}

\author{
\alignauthor{}
Dan Stelljes\\
  \affaddr{Division of Science and Mathematics}\\
  \affaddr{University of Minnesota, Morris}\\
  \affaddr{Morris, Minnesota, USA 56267}\\
  \email{stell124@morris.umn.edu}
}
\conferenceinfo{UMM CSci Senior Seminar Conference, December 2016}{Morris, MN}
\title{Composable Concurrency Models}

@@

\begin{abstract}

The need to manage concurrent operations in applications has led to the development of a wide variety of concurrency models. Modern programming languages each provide several concurrency models to serve different requirements, and programmers benefit from being able to use them in tandem. We discuss challenges surrounding concurrency control and examine situations in which conflicts between models can occur.

\end{abstract}

@@

Explanation of why concurrency is desirable (and usually necessary in practice). Example of an application that relies heavily on concurrency (web browser, for instance).

\section{Background}

\subsection{Concurrency}

Brief introduction to threads, processes, and cores. Possible discussion of concurrency primitives (coroutines, fibers, green threads).

\subsection{Atomicity}

Explanation of individual effects/mutual exclusion.

@@

Explanation of execution order as a requirement for correctness.

\subsection{Visibility}

Explanation of thread opacity and communication.

\section{Common concurrency models}

\subsection{Atomic variables}

Atomic variables can only be read and mutated by operations that guarantee atomicity. For example, the \emph{compare-and-swap} operation compares the current value of the variable to a given value and only performs a write if the values match:

\begin{lstlisting}[language=C]
  bool swap(int* v, int old, int new) {
    if (*v != old) {
      return false;
    }

    *v = new;
    return true;
  }
\end{lstlisting}

\emph{compare-and-swap} ensures that a new value cannot be based on outdated information. Suppose that a thread \emph{A} reads a variable \emph{v} and begins computing a new value. At the same time, another thread \emph{B} modifies \emph{v}. When \emph{A} tries to set the value of \emph{v} via \emph{compare-and-swap}, the write will fail. As a result, race conditions are prevented and all writes on \emph{v} are guaranteed to be atomic.

Languages provide varying levels of support for atomic variables. Go has no atomic variables, but implements some low-level atomic operations (including \emph{compare-and-swap}). Java offers atomic analogues to its primitive wrapper classes (e.g., \code{AtomicInteger} as an atomic \code{Integer}) that ensure thread safety and allow implementations to take advantage of hardware-specific atomic instructions. Clojure provides \code{atom} as a reference type; the value of an \code{atom} can be read with \code{deref} and written with \code{swap!}.

\subsection{Software transactional memory}

Software transactional memory (STM) allows multiple concurrent operations to transactionally write to a shared location in memory. STM is an example of optimistic concurrency control: Each operation writes to the shared memory with no regard to the activity of other threads. After the entire transaction is completed, the transaction manager verifies that other threads have not also made changes to the shared memory. If there are conflicting changes, the transaction is aborted and re-executed until it eventually succeeds.~\cite{Shavit1995}

Conceptually, STM simplifies concurrency because it allows transactions to be thought of as a single-threaded operation. A thread cannot observe changes made to other threads while a transaction is in progress, nor can other threads observe modifications by that thread until the transaction is completed. Only when a transaction completes successfully will changes will become visible to other threads.

Harris and Fraser proposed the use of critical regions to represent transactions:~\cite{Harris2014}

\begin{lstlisting}
  atomic (index > 0) {
    index--;
    return buffer[index];
  }
\end{lstlisting}

In this (somewhat contrived) example, everything in the \code{atomic} block is executed as a transaction. The optional guard condition \code{index > 0} prevents a transaction from occurring if it would fail or is unnecessary. No popular languages currently support this type of representation; most often, STM is implemented by a library. The Concurrent Ruby library, for example, models transactions with \code{TVar} and \code{Concurrent::atomically}:~\cite{ConcurrentRubyTVar}

\begin{lstlisting}
  v = TVar.new(50)

  Concurrent::atomically do
    v.value += 50
  end
\end{lstlisting}

\subsection{Communicating threads}

Instead of sharing memory, some concurrency models restrict operations to private memory. Threads communicate strictly by message passing, which avoids race conditions. Communicating sequential processes (CSP), one of the oldest communicating thread models, describes systems in terms of independent processes that communicate through predefined channels.~\cite{Hoare1978} Other models, most notably the actor model, rely on asynchronous message passing to specific entities.~\cite{Agha1986}

Clojure is useful as a demonstration of communicating threads in that it provides several different models.~\cite{Swalens2014} CSP is implemented by the core.async library. \code{chan} opens a new channel and \code{go} creates a new thread. Inside a \code{go} block, values can be added to and removed from a channel; subsequent operations will block until another thread performs a complementary operation.

Clojure's agents, by contrast, are asynchronous---there are no blocking receive operations. An agent represents the state of an individual entity and only allows changes to occur as the result of an action. Actions are applied to the state of an agent and return the new state of the agent.

The actor model is not implemented by Clojure, though it is commonly used in distributed systems. (The ``pure'' use of actors assumes that everything is an actor.) Actors can only communicate via asynchronous message passing. Upon receiving a message, an actor can choose to send messages to other actors, create new actors, and determine the behavior used for the next received message.

\subsection{Proxies}

Proxies are placeholders for values that are the result of some concurrently executed operation. Futures and promises are the two main types of proxy, though the terms are frequently used interchangably (along with ``delayed'' and ``deferred''). Specifically, futures are resolved to the result of the completed operation. The result is then accessed implicitly; any use of the future will return its value. Promises can be created independently and require the result to be accessed explicitly.

JavaScript relies heavily on promises to model concurrency without descending into ``callback hell.'' Rather than deeply nesting callback functions, asynchronous actions can be expressed as promises:

\begin{lstlisting}
  var read = new Promise((resolve) => {
    fs.readdir('.', resolve);
  });

  var count = (files) => {
    return new Promise((resolve) {
      resolve(files.length);
    });
  };

  var print = console.log;

  read.then(count).then(print);
\end{lstlisting}

\section{Correctness criteria}

\subsection{Safety}

Definition of safety as a guarantee of partial correctness (``nothing bad will happen'').

\subsection{Liveness}

Definition of liveness as a guarantee of progress and eventual termination (``something good will eventually happen'').

\section{Composability challenges}

Atomics: Operations affecting multiple atomics may cause race conditions. ABA problem.

STM: No irrevocable operations can be performed.

Communicating threads: CSP relies on blocking for communication, and deadlocks are almost inevitable when combined with other models.

Proxies: Reading future/promise values before they're actually available can cause deadlocks. Also, mutually recursive futures.
